---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)
```

# denim <img src="man/figures/logo.svg" align="right" width="120" />

<!-- badges: start -->
[![Project Status: Active – The project has reached a stable, usable state and is being actively developed.](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active)
[![R-CMD-check](https://github.com/thinhong/denim/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/thinhong/denim/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/thinhong/denim/branch/master/graph/badge.svg)](https://app.codecov.io/gh/thinhong/denim?branch=master)
<!-- badges: end -->

An R package for building and simulating deterministic discrete-time compartmental models with memory.

## Installation

You can install `denim` from [CRAN](https://cran.r-project.org) with:

``` r
install.packages("denim")
```

You can install the development version of denim from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("thinhong/denim")
```

## Example

This is a basic example to illustrate the specification of a simple SIR epidemiological model, which contains three compartments: susceptible (S), infected (I) and recovered (R). The durations spent in the I compartment are gamma distributed and this is specified on the `I -> R` transition:

```{r example}
library(denim)

transitions <- list(
  "S -> I" = "beta * S * I / N",
  "I -> R" = d_gamma(3, 2)
)

parameters <- c(
  beta = 0.12,
  N = 1000
)

initialValues <- c(
  S = 999, 
  I = 1, 
  R = 0
)

simulationDuration <- 10
timeStep <- 0.01

mod <- sim(transitions = transitions, initialValues = initialValues, 
           parameters = parameters, simulationDuration = simulationDuration, 
           timeStep = timeStep)
```

The output is a data frame with 4 columns: `Time`, `S`, `I` and `R`

```{r}
head(mod)
```

We can plot the output with:

```{r example-plot, dpi=300, fig.width=5}
plot(mod)
```
