---
title: "denim benchmark"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{denim benchmark}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
python:
  python_path: /opt/anaconda3/envs/bnn/bin/python
bibliography: references.bib
---

To assess denim's performance, we benchmark it against other approaches and packages.

The benchmarking process was conducted on a Macbook with M2 Pro chip, 16 GBs of RAM and 10 cores (6 performance and 4 efficiency). 

## Benchmark settings

All approaches will simulate the following SEIR model, with the same simulation duration of 180

```{r echo=FALSE}
DiagrammeR::grViz("digraph {
  graph [layout = dot, rankdir = LR]
  
  node [shape = rectangle]
  
  S -> E [label = '(R0/tr) * S * (I / N) * td']
  E -> I [label = 'd_gamma(1/4,2)']
  I -> R [label = 'd_gamma(1/3,2)']
  }",
  width = 700, height = "100%")
```

Each approach will be run 50 times

```{r benchmark settings}
total_runs <- 50L # number of runs
sim_duration <- 180 # duration of simulation
```

## uSEIR

Simulate model using uSEIR approach [@Hernndez2021]

Source code: <https://github.com/jjgomezcadenas/useirn/blob/master/nb/uSEIR.ipynb>

<details>

<summary>load useir implementation in pure Python</summary>

```{r}
library(reticulate)
# use_python("/opt/anaconda3/envs/bnn/bin/python", required = TRUE)
use_condaenv(condaenv='bnn', required = TRUE)
matplotlib <- import("matplotlib")
matplotlib$use("Agg", force = TRUE)
py_run_file("../supplements/useir_python.py")
```

</details>

### Python implementation

<details>

<summary>Code for running uSEIR in pure Python</summary>

```{python}
#| output: false

import time
import concurrent.futures
import pickle
import os
from statistics import mean

python_runs = []

def get_python_runtime(n):
  start = time.time()
  df = solve_uSeir(ti_shape     = 2,
                     ti_scale     = 4,
                     tr_shape     = 2,
                     tr_scale     = 3,
                     R0           = 3.5)
  end = time.time()
  return end - start
  
# load cached result if available instead of rerun due to long run time
cached_python_runs = "../supplements/python_runs.pkl"
if os.path.exists(cached_python_runs):
  # If the file exists, load the Python list from the file
  with open(cached_python_runs, 'rb') as f:
    python_runs = pickle.load(f)
else:
  print("no cache found")
  # multithread instead for quicker result
  with concurrent.futures.ProcessPoolExecutor(max_workers=8) as executor:
    python_runs = list(executor.map(get_python_runtime, range(r.total_runs)))
  
# mean(python_runs)

# plot_useir((df,), ('G',), T = 'uSEIR', figsize=(14,8))
# print(f'python solve_seir call: dr = {end-start}')
```

</details>

Run time for uSEIR approach, Python implementation (in seconds)

```{r}
py$python_runs
```

Median run time for uSEIR approach, Python implementation: `r median(py$python_runs)` seconds

### Cython implementation

<details>

<summary>Code for running uSEIR in Cython (C backend)</summary>

```{python}
#| output: false

# import precompiled cython module
import sys
sys.path.insert(0, "../supplements")
import useir
import time
import pyarrow as pa

cython_runs = []

# --- Get runtime ----
for i in range(r.total_runs):
  start = time.time()
  df = useir.csolve_uSeir(dist = "gamma",
                    ti_shape     = 2,
                     ti_scale     = 4,
                     tr_shape     = 2,
                     tr_scale     = 3,
                     R0           = 3.5
  )
  end = time.time()

  cython_runs = cython_runs + [end - start]

# ---- Get output for uSEIR -----
df = useir.csolve_uSeir(dist = "gamma",
                    ti_shape     = 2,
                     ti_scale     = 4,
                     tr_shape     = 2,
                     tr_scale     = 3,
                     R0           = 3.5
  )
  
# convert to pyarrow table for easy conversion to R data.frames
to_r_df = pa.Table.from_pandas(df)
```

</details>

Run time for uSEIR approach, Cython implementation (in seconds)

```{r}
py$cython_runs
```

Median run time for uSEIR approach, Cython implementation: `r median(py$cython_runs)` seconds

## deSolve

### Model in R

```{=html}
<details>
  <summary>Code for running SEIR in <code>deSolve</code></summary>
```

```{r}
library(deSolve)
parameters <- c(gamma_rate_I = 1/4, shape_I=2,
                gamma_rate_R = 1/3, shape_R = 2,
                R0 = 3.5, N = 1e6) 
initialValues <- c(S = 999999, E1 = 1,
                   E2 = 0, E = 0, I1=0, 
                   I2=0, I=0, R=0
                   )

# --- Transition def for deSolve
transition_func <- function(t, state, param){
  with(as.list( c(state, param) ), {
      tr = shape_R*(1/gamma_rate_R)
      
      dS = - (R0/tr) * S * I/N
      # apply linear chain trick
      dE1 = (R0/tr) * S * I/N - gamma_rate_I*E1
      dE2 = gamma_rate_I*E1 - gamma_rate_I*E2
      dE = dE1 + dE2
      dI1 = gamma_rate_I*E2 - gamma_rate_R*I1
      dI2 = gamma_rate_R*I1 - gamma_rate_R*I2
      dI =  dI1 + dI2

      dR = gamma_rate_R*I2
      list(c(dS, dE1, dE2, dE, dI1, dI2, dI, dR))
  })
}

times <- seq(0, sim_duration, 1)

# ------ Compute run time ------
desolve_runs <- bench::mark(
  ode(y = initialValues, times = times, parms = parameters, func = transition_func),
  iterations = total_runs
)

ode_mod <- ode(y = initialValues, times = times, parms = parameters, func = transition_func)

ode_mod <- as.data.frame(ode_mod)
```

</details>

Run time for `deSolve` implementation

```{r}
desolve_runs$time
```
Median run time for deSolve, with model defined in R: `r desolve_runs$median` seconds

### Model in C

```{=html}
<details>
  <summary>Code for running model defined in C</summary>
```
```{r}
# compile model
# system("R CMD SHLIB supplements/desolve_mod/benchmark_mod.c")

# compiled file on Windows will have .dll extension instead of .so
dyn.load("../supplements/desolve_mod/benchmark_mod.so")

initialValues <- c(S = 999999, E1 = 1,
                   E2 = 0, E = 0, I1=0, 
                   I2=0, I=0, R=0
                   )

parameters <- c(R0 = 3.5, scale_I = 4, shape_I=2,
                scale_R = 3, shape_R = 2, N = 1e6) 

deSolve_c_runs <- bench::mark(
  # run model defined in C
  ode(initialValues, times, func = "derivs", parms = parameters,
  dllname = "benchmark_mod", initfunc = "initmod"),
  iterations = total_runs
)

dyn.unload("../supplements/desolve_mod/benchmark_mod.so")
```

</details>

Run time for `deSolve` with model defined in C

```{r}
deSolve_c_runs$time
```
Median run time for deSolve, with model defined in C: `r deSolve_c_runs$median` seconds

### Model in Fortran

```{=html}
<details>
  <summary>Code for running model defined in C</summary>
```
```{r}
# compile model in fortran
# system("R CMD SHLIB supplements/desolve_mod/benchmark_mod_fortran.f")

dyn.load("../supplements/desolve_mod/benchmark_mod_fortran.so")

initialValues <- c(S = 999999, E1 = 1,
                   E2 = 0, E = 0, I1=0, 
                   I2=0, I=0, R=0
                   )

parameters <- c(R0 = 3.5, scale_I = 4, shape_I=2,
                scale_R = 3, shape_R = 2, N = 1e6) 

deSolve_fortran_runs <- bench::mark(
  # run model defined in C
  ode(initialValues, times, func = "derivs", parms = parameters,
  dllname = "benchmark_mod_fortran", initfunc = "initmod"),
  iterations = total_runs
)

dyn.unload("../supplements/desolve_mod/benchmark_mod_fortran.so")
```

```{=html}
</details>
```
```{r}
deSolve_fortran_runs$time
```
Median run time for deSolve, with model defined in Fortran: `r deSolve_fortran_runs$median` seconds

## denim

<details>

<summary>Code for running SEIR in `denim`</summary>

```{r}
timeStep <- 0.01

library(denim)

denim_model <- list(
  "S -> E" = "(R0/tr) * timeStep * S * (I/N)", # formulate according that of uSEIR method
  "E -> I" = d_gamma(rate = 1/4, shape = 2),
  "I -> R" = d_gamma(rate = 1/3, shape = 2)
)

initialValues <- c(S = 999999, E = 1, I= 0, R= 0)
parameters <- c(R0 = 3.5, 
                tr = 3*2, # compute mean recovery time, for gamma it's scale*shape
                N = 1e6)

# ---- Get runtimes ----
denim_runs <- bench::mark(
  sim(transitions = denim_model,
                     initialValues = initialValues,
                     parameters = parameters,
                     simulationDuration = sim_duration, timeStep = timeStep),
  iterations = total_runs
)


# ---- Get output ----
denim_out <- sim(transitions = denim_model, 
                     initialValues = initialValues,
                     parameters = parameters,
                     simulationDuration = sim_duration, timeStep = timeStep)
```

</details>

Run time for `denim` implementation

```{r}
denim_runs$time
```

Median run time for denim: `r denim_runs$median` seconds

## Visualize output

```{r echo=FALSE, warning=FALSE, message=FALSE}
library(arrow)
library(reticulate)

df <- as.data.frame(py$to_r_df)

# comput prop to compare with uSEIR
rescaled_denim <- denim_out
rescaled_denim[, c("S","E", "I", "R")] <- denim_out[, c("S","E", "I", "R")]/1e6

# compute prop for deSolve output as well
rescaled_desolve <- ode_mod
rescaled_desolve[, c("S","E", "I", "R")] <- rescaled_desolve[, c("S","E", "I", "R")]/1e6

useir_offset <- 6

plot_seir_util <- function(df, mod_name="denim", time_col = "Time", 
                           ylab = "Proportion",
                           x_legend=150, y_legend=0.7){
  plot(x = df[[time_col]], y = df[["S"]],
       xlab = "time", ylab = "Prop", ylim = c(0, 1),
       main=paste0(mod_name, " output"), col = "#4876ff", type="l", lwd=2)
  lines(df[[time_col]], df[["E"]], lwd=2, col = "blueviolet")
  lines(df[[time_col]], df[["I"]], lwd=2, col = "red")
  lines(df[[time_col]], df[["R"]], lwd=2, col = "green")
  legend(x = x_legend, y = y_legend,
         legend=c("S", "E", "I", "R"),
         col = c("#4876ff", "blueviolet", "red", "green"),
         lty = 1)
}

# --- denim output -----
plot_seir_util(rescaled_denim)

# --- deSolve output ----
plot_seir_util(rescaled_desolve, time_col = "time", mod_name = "deSolve")

# --- uSEIR output ---- 
plot_seir_util(df, time_col = "t", mod_name = "uSEIR")
```

## Compare run time

The following plot shows run time for 50 runs (with horizontal line showing median run time) of each approach.

```{r echo=FALSE}
par(xpd=T, mar=par()$mar+c(0,0,0,6.5))

plot(x = 1:total_runs, y = py$cython_runs, type = "p", col = "blueviolet",
     ylim = c(0,1.2),
     main = paste0("Run time for ", total_runs, " runs"),
     xlab = "Run", ylab = "Run time (in seconds)")
lines(x = 1:total_runs, y = rep(median(py$cython_runs), total_runs), col = "blueviolet", lty = 2)
points(x = 1:total_runs, y = as.numeric(denim_runs$time[[1]]), col = "red")
lines(x = 1:total_runs, y = rep(median(denim_runs$time[[1]]), total_runs), col = "red", lty = 2)
points(x = 1:total_runs, y = as.numeric(desolve_runs$time[[1]]), col = "darkgreen")
lines(x = 1:total_runs, y = rep(median(desolve_runs$time[[1]]), total_runs), col = "darkgreen", lty = 2)
legend(x = 53, y = 0.9,
       legend = c("uSEIR (Cython)", "denim", "deSolve (model in R)"), 
       col = c("blueviolet", "red", "darkgreen"),
       lty = 1, cex = 0.7)

# Restore default clipping rect
par(mar=c(5, 4, 4, 2) + 0.1)
```


## Run time scaling in denim
It is worth noting that runtime for denim is also dependent on duration of time step (`timeStep` parameter for `sim`). 

The following plot demonstrates how run time changes as value for timeStep changes, using the same model for benchmarking. The values for timeStep being evaluated are `[0.01, 0.02, 0.05, 0.1, 0.25, 0.5, 1]`.

```{r echo=FALSE}
timeStepDurations <- c(0.01, 0.02, 0.05, 0.1, 0.25, 0.5, 1)
runtime_scaling <- lapply(timeStepDurations, 
     \(x){
       time <- bench::mark(
         sim(transitions = denim_model,
                   initialValues = initialValues,
                   parameters = parameters,
                   simulationDuration = sim_duration, timeStep = x)
       )
       
       time$median
     }
   )
```
```{r echo=FALSE}
plot(x = timeStepDurations, y = runtime_scaling, col = "cornflowerblue",
     xlab = "Time step duration",
     ylab = "Median run time (in seconds)")
lines(x = timeStepDurations, y = runtime_scaling, col = "cornflowerblue",
     xlab = "Time step duration",
     ylab = "Median run time (in seconds)")
abline(v = timeStep, col = "red", lty = 2, lwd = 2)  
abline(v = 1, col = "blueviolet", lty = 2, lwd = 2)  
text(x = timeStep, y = max(unlist(runtime_scaling)), labels = "timeStep for benchmarking", col = "red", pos = 4)
text(x = 1, y = max(unlist(runtime_scaling)), labels = "default timeStep", col = "blueviolet", pos = 2)
```

