---
title: "denim vs deSolve"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{denim vs deSolve}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Comparison between deSolve and denim

Using `deSolve`, we can replicate **Erlang distribution** and **exponential distribution** for testing

```{r define variables}
library(denim)
library(deSolve)

# --- Transition def for denim
transitions <- list(
  "S -> I" = d_exponential(0.2),
  "I -> R" = d_gamma(3, 2)
)
parameters <- c(rate = 0.2, scale = 3, shape=2) 
initialValues <- c(S = 999, I = 1, I1 = 1, I2=0, R=0)

# --- Transition def for deSolve
transition_func <- function(t, state, param){
  with(as.list( c(state, param) ), {
      gamma_rate = 1/scale
      dS = -rate*S
      # apply linear chain trick
      dI1 = rate*S - gamma_rate*I1
      dI2 = gamma_rate*I1 - gamma_rate*I2
      dI =  dI1 + dI2
      dR = gamma_rate*I2
      list(c(dS, dI, dI1, dI2, dR))
  })
}

# --- Timestep definition
simulationDuration <- 20 
timestep <- 0.01 # small timestep required for comparison
```

### Run simulation with denim

```{r}
denim_start <- Sys.time()
mod <- sim(transitions = transitions, initialValues = initialValues, parameters, simulationDuration = simulationDuration, timeStep = timestep)
denim_end <- Sys.time()

# --- show output
head(mod[mod$Time %in% 1:simulationDuration,])
```

### Run simulation with deSolve

```{r}
times <- seq(0, simulationDuration, timestep)

desolve_start <- Sys.time()
ode_mod <- ode(y = initialValues, times = times, parms = parameters, func = transition_func)
desolve_end <- Sys.time()

# --- show output
ode_mod <- as.data.frame(ode_mod)
head(ode_mod[ode_mod$time %in% 1:simulationDuration, c("time", "S", "I", "R")])
```

### Execution time comparison

```{r include=FALSE}
time_diff <- round(as.numeric(difftime(denim_end, denim_start, units="sec")) /
                     as.numeric(difftime(desolve_end, desolve_start, units="sec")), digits = 2)
```

`denim` takes approximately `r time_diff` times as long as `deSolve` to compute the result with the given specifications .

This significant difference can be attributed to the difference in approaches: `deSolve` solves a system of ODEs while `denim` iterates through each timestep and updates the population in each compartment

While the approach in `denim` allow more flexibility in types of dwell time distributions, the computation time scales up as timestep grows smaller (in `O(n)` time complexity).

## Plot the result

```{r}
# increase timestep before plotting
mod <- mod[mod$Time %in% seq(0, simulationDuration, 0.2),]
ode_mod <- ode_mod[ode_mod$time %in% seq(0, simulationDuration, 0.2),]
```

```{r,echo=FALSE, fig.width=8, fig.height=5}
# ---- Plot S compartment
plot(x = mod$Time, y = mod$S,xlab = "Time", ylab = "Count", main="S compartment",
     col = "#4876ff", type="l", lwd=3)
lines(ode_mod$time, ode_mod$S, lwd=3, lty=3)
legend(x = 15, y = 900,legend=c("denim", "deSolve"), col = c("#4876ff", "black"), lty=c(1,3))

# ---- Plot I compartment
plot(x = mod$Time, y = mod$I, xlab = "Time", ylab = "Count", main="I compartment",
      col = "#4876ff", type="l", lwd=2)
lines(ode_mod$time, ode_mod$I, lwd=3, lty=3)
legend(x = 15, y = 350,legend=c("denim", "deSolve"), col = c("#4876ff", "black"), lty=c(1,3))

# ---- Plot R compartment
plot(x = mod$Time, y = mod$R, xlab = "Time", ylab = "Count", main="R compartment",
     col = "#4876ff", type="l", lwd=2)
lines(ode_mod$time, ode_mod$R, lwd=3, lty=3)
legend(x = 15, y = 300,legend=c("denim", "deSolve"), col = c("#4876ff", "black"), lty=c(1,3))

```

## Comparison with SEIR model

### Implementation using deSolve

<details>
  <summary> model definition and runtime using `deSolve` </summary>
```{r }
library(deSolve)
parameters <- c(scale_I = 4, shape_I=2,
                scale_R = 3, shape_R = 2,
                timeStepDur = 1, R0 = 3.5, N = 1e6) 
initialValues <- c(S = 999999, E1 = 1,
                   E2 = 0, E = 0, I1=0, 
                   I2=0, I=0, R=0
                   )

# --- Transition def for deSolve
transition_func <- function(t, state, param){
  with(as.list( c(state, param) ), {
      gamma_rate_I = 1/scale_I
      gamma_rate_R = 1/scale_R
      
      tr = scale_R*shape_R
      
      dS = - (R0/tr) * S * I/N
      # apply linear chain trick
      dE1 = (R0/tr) * S * I/N - gamma_rate_I*E1
      dE2 = gamma_rate_I*E1 - gamma_rate_I*E2
      dE = dE1 + dE2
      dI1 = gamma_rate_I*E2 - gamma_rate_R*I1
      dI2 = gamma_rate_R*I1 - gamma_rate_R*I2
      dI =  dI1 + dI2

      dR = gamma_rate_R*I2
      list(c(dS, dE1, dE2, dE, dI1, dI2, dI, dR))
  })
}

times <- seq(0, 210, 1)

# ------ Compute mean for 10 runs ------
# runtimes <- sapply(1:10,
#   function(i){
#     time <- system.time(
#       ode(y = initialValues, times = times, parms = parameters, func = transition_func)
#     )
#     time[["elapsed"]]
#     
#   }
#   
# )
# mean(runtimes)

ode_mod <- ode(y = initialValues, times = times, parms = parameters, func = transition_func)

# --- show output
ode_mod <- as.data.frame(ode_mod)
```

</details>

Runtime for deSolve is `~ 0.013`

### Implementation using denim

<details>
  <summary> model definition and runtime using `denim` </summary>

```{r }
denim_model <- list(
  "S -> E" = "(R0/tr) * timeStepDur * S * (I/N)", # formulate according that of uSEIR method
  "E -> I" = d_gamma(scale = 4, shape = 2),
  "I -> R" = d_gamma(scale = 3, shape = 2)
)

initialValues <- c(S = 999999, E = 1, I= 0, R= 0)
parameters <- c(R0 = 3.5, 
                tr = 3*2, # compute mean recovery time, for gamma it's scale*shape
                N = 1e6, timeStepDur = 0.01)

# ------ Compute mean for 10 runs ------
# runtimes <- sapply(1:10,
#   function (i) {
#     time <- system.time(
#       sim(transitions = denim_model, 
#                      initialValues = initialValues,
#                      parameters = parameters,
#                      simulationDuration = 210, timeStep = 0.01)
#     )
#     return(time[["elapsed"]])
#   })
# mean(runtimes)

mod <- sim(transitions = denim_model, 
                     initialValues = initialValues,
                     parameters = parameters,
                     simulationDuration = 210, timeStep = 0.01)

# denim_out[, c("S","E", "I", "R")] <- denim_out[, c("S","E", "I", "R")]/1e6
plot(mod)
```

</details>

Runtime for denim is `~ 1.0548`


### Plot result

```{r, echo=FALSE, fig.width=8, fig.height=5}
# ---- Plot S compartment
plot(x = mod$Time, y = mod$S,xlab = "Time", ylab = "Count", main="S compartment",
     col = "#4876ff", type="l", lwd=3)
lines(ode_mod$time, ode_mod$S, lwd=3, lty=3)
legend(x = 15, y = 4e5,legend=c("denim", "deSolve"), col = c("#4876ff", "black"), lty=c(1,3))

# ---- Plot E compartment
plot(x = mod$Time, y = mod$E, xlab = "Time", ylab = "Count", main="E compartment",
      col = "#4876ff", type="l", lwd=2)
lines(ode_mod$time, ode_mod$E, lwd=3, lty=3)
legend(x = 15, y = 1e5,legend=c("denim", "deSolve"), col = c("#4876ff", "black"), lty=c(1,3))

# ---- Plot I compartment
plot(x = mod$Time, y = mod$I, xlab = "Time", ylab = "Count", main="I compartment",
      col = "#4876ff", type="l", lwd=2)
lines(ode_mod$time, ode_mod$I, lwd=3, lty=3)
legend(x = 15, y = 1e5,legend=c("denim", "deSolve"), col = c("#4876ff", "black"), lty=c(1,3))

# ---- Plot R compartment
plot(x = mod$Time, y = mod$R, xlab = "Time", ylab = "Count", main="R compartment",
     col = "#4876ff", type="l", lwd=2)
lines(ode_mod$time, ode_mod$R, lwd=3, lty=3)
legend(x = 15, y = 4e5,legend=c("denim", "deSolve"), col = c("#4876ff", "black"), lty=c(1,3))
```
