---
title: "denim: deterministic discrete-time model with memory"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{denim: deterministic discrete-time model with memory}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align="center"
)
library(DiagrammeR) # for flowchart diagram
library(denim)
```

## 1. Simple SIR model with gamma distributed lengths of stay

The SIR model uses 3 compartments: S (susceptible), I (infected), R (recovered) to describe clinical status of individuals. We use the most simple form of SIR model to demonstrate how to define the distribution of the lengths of stay distribution.

<center>

```{r, echo=FALSE}
DiagrammeR::grViz("digraph {
  graph [layout = dot, rankdir = LR]
  
  node [shape = rectangle]
  
  S -> I [label = '&#946;SI/N']
  I -> R [label = '&#947;I']
  }",
  width = 300, height = "100%")
```

</center>

The model equations are:

$$S_{t+1} - S_{t} = -\lambda S_{t} = -\frac{\beta I_{t}}{N}S_{t}$$ $$I_{t+1} - I_{t} = \frac{\beta I_{t}}{N}S_{t} - \gamma I_{t}$$ $$R_{t+1} - R_{t} = \gamma I_{t}$$

-   $N$: total population size, $N = S + I + R$
-   $\beta$: the product of contact rates and transmission probability; usually we define $\lambda =\frac{\beta I_{t}}{N}$ as the force of infection
-   $\gamma$: recovery rate

Usually to solve the model easier we make an assumption that the recovery rate $\gamma$ is constant, this will leads to an exponentially distributed length of stay i.e most individuals recover after 1 day being infected.

```{r, echo=FALSE}
rates <- c(0.5, 1, 1.5)
x <- seq(0, 5, 0.001)
y <- dexp(x = x, rate = rates[1])
y2 <- dexp(x = x, rate = rates[2])
y3 <- dexp(x = x, rate = rates[3])

col_codes <- c("#374F77", "#EF9F32", "#6ECBD7")
plot(x, y, type = "l", col = col_codes[1], lty = 1, lwd = 3,
     xlab = "Length of stay (days)", ylab = "", 
     ylim = c(0, 1.5), yaxt = 'n')
lines(x, y2, col = col_codes[2], lty = 1, lwd = 3)
lines(x, y3, col = col_codes[3], lty = 1, lwd = 3)
legend("right", legend = c(0.5, 1.0, 1.5), 
       col = col_codes, lty = 1, lwd = 3, bty = "\n")
```

A more realistic length of stay distribution can look like this, of which most patients recovered after 4 days. We defined this using a gamma distribution with shape = 3 and rate = 1/2.

```{r, echo=FALSE}
x <- seq(0, 20, 0.001)
y <- dgamma(x = x, shape = 3, rate = 1/2)

plot(x, y, type = "l", col = col_codes[1], lty = 1, lwd = 3,
     xlab = "Length of stay (days)", ylab = "", yaxt = 'n')
```

The model now look like this:

<center>

```{r, echo=FALSE}
DiagrammeR::grViz("digraph {
  graph [layout = dot, rankdir = LR]
  
  node [shape = rectangle]
  
  S -> I [label = '&#946;SI/N']
  I -> R [label = 'd_gamma(1/2, 3)']
  }",
  width = 400, height = "100%")
```

</center>

**Model specification**

*Model transition*

<center>

```{r, echo=FALSE}
DiagrammeR::grViz("digraph {
  graph [layout = dot, rankdir = LR]
  
  node [shape = rectangle]
  
  S -> I [label = '&#946;SI/N']
  I -> R [label = 'd_gamma(1/2, 3)']
  }",
  width = 400, height = "100%")
```

</center>

We have two transitions `S -> I` and `I -> R` in this case. The transitions are specified in a list follow this format `"transition" = equation`, of which equation is defined with one of our functions for waiting time distribution.

```{r}
transitions <- list(
  "S -> I" = "beta * S * I / N",
  "I -> R" = d_gamma(rate = 1/2, 3)
)
```

*Initial state*

Use a vector to define the compartments with their assigned names and initial values in the format `compartment_name = initial_value`:

```{r}
initialValues <- c(
  S = 999, 
  I = 1, 
  R = 0
)
```

*Model parameters*

If we use a math expression, any symbols except the compartment names are parameters, and would be defined by constant values. There are two constant parameters in our example: `beta` and `N`:

```{r}
parameters <- c(
  beta = 0.012,
  N = 1000
)
```

**Model application**

*Time step specification*

We run the model for 30 days and give output at 0.01 daily intervals. The default interval (time step) is 1 if not declared explicitly.

```{r}
simulationDuration <- 30
timeStep <- 0.01
```

```{r, fig.width = 6}
mod <- sim(transitions = transitions, 
           initialValues = initialValues, 
           parameters = parameters, 
           simulationDuration = simulationDuration, 
           timeStep = timeStep)
head(mod)
plot(mod)
```

## 2. How the algorithm work?

In the SIR model, all infected individuals are presented by a single compartment I and have the same recovery rate $\gamma$.

<center>

```{r, echo=FALSE}
DiagrammeR::grViz("digraph {
  graph [layout = dot, rankdir = LR]
  
  node [shape = rectangle]
  
  S -> I [label = '&#946;SI/N']
  I -> R [label = '&#947;I']
  }",
  width = 300, height = "100%")
```

</center>

We want the recovery rate of individuals who had been infected for 1 day differ from the recovery rate of 2-day infected patients. So rather than using one compartment for infected (I), we define multiple infected sub-compartments. The number of sub-compartments depends on the maximum day we expect all infected individuals would be recovered.

For example, if we expect a disease with a maximum 4 days of infection, we will end up with 4 sub-compartments. Each sub-compartment has its own recovery rate $\gamma_{1}$, $\gamma_{2}$, $\gamma_{3}$, $\gamma_{4}$. At day 4 it is certain that the patient will recover (because we assume that this disease has a maximum 4 days of infection), $\gamma_{4} = 1$.

<center>

```{r, echo=FALSE}
DiagrammeR::grViz("digraph {
  graph [layout = dot, rankdir = LR]
  
  node [shape = rectangle]
  
  S -> I1 [label = '&#946;S(I@_{1} + I@_{2} + I@_{3} + I@_{4})/N']
  I1 -> R1 [label = '&#947;@_{1}I@_{1}']
  I2 -> R2 [label = '&#947;@_{2}I@_{2}']
  I3 -> R3 [label = '&#947;@_{3}I@_{3}']
  I4 -> R4 [label = 'I@_{4}']
  I1 -> I2 [label = '(1-&#947;@_{1})I@_{1}']
  I2 -> I3 [label = '(1-&#947;@_{2})I@_{2}']
  I3 -> I4 [label = '(1-&#947;@_{3})I@_{3}']
  }", height = "100%", width = "100%")
```

</center>

Let $R_1 + R_2 + R_3 + R_4 = \Sigma R$. We have $\frac{R_1}{\Sigma R} = p_1$, $\frac{R_2}{\Sigma R} = p_2$, $\frac{R_3}{\Sigma R} = p_3$, $\frac{R_4}{\Sigma R} = p_4$. Our mission is to estimate $\gamma_{1}$, $\gamma_{2}$, $\gamma_{3}$ to obtain $p_1$, $p_2$, $p_3$, $p_4$ that fit a pre-defined distribution at the equilibrium state. This can be obtained by setting:

$$\gamma_{i} = \frac{p_i}{1 - \sum_{j=1}^{i-1}p_j}$$

For a given length of stay distribution, we identify the maximum length of stay using its cumulative distribution function. Because cumulative distribution function is asymptotic to 1 and never equal to 1, we need to set a value that is acceptable to be rounded to 1. If we want a cumulative probability of 0.999 to be rounded as 1, we set the error tolerance threshold as `1 - 0.999 = 0.001` (specified by the argument `errorTolerance = 0.001`). The time when cumulative probability = 0.999 will be set as the maximum length of stay of the compartment. Default `errorTolerance` of `denim` is set at `0.001`.

*Initialize population in sub-compartments*

By default, the initial population is always assigned to the first sub-compartment (for example, if the initial value for I compartment is `I = 1`, denim will initialize `I1 = 1` while `I2 = I3 = I4 = 0`).

User can also choose to distribute initial population across sub-compartments based on the specified distribution, i.e. with initial population `I = n`, then denim will initialize `I1 = n*p1`, `I2 = n*p2`, `I3 = n*p3`, `I4 = n*p4`. To set up the initial population in this way, simply specify the parameter `dist_init = TRUE` (only applicable for distribution transitions).

```{r eval=FALSE}
transitions <- list(
  "S -> I" = "beta * S * I / N",
  "I -> R" = d_gamma(rate = 1/2, 3, dist_init=TRUE)
)
```

## 3. Waiting time distribution

`denim` offers 2 main ways to define a transition: either by a waiting time distribution, or a mathematical expression.

Current available distributions in this package including:

-   `d_exponential(rate)`: Discrete **exponential distribution** with parameter `rate`

-   `d_gamma(rate, shape)`: Discrete **gamma distribution** with parameters `rate` and `shape`

-   `d_weibull(scale, shape)`: Discrete **Weibull distribution** with parameters `scale` and `shape`

-   `d_lognormal(mu, sigma)`: Discrete **log-normal distribution** with parameters `mu` and `sigma`

-   `nonparametric(waitingTimes...)`: A **vector of values**, could be numbers, percentages, density of the length of stay based on real data, `denim` will convert it into a distribution

Mathematical expression: Transition defined with a string value such as `"beta * S * I / N"` will be converted to a mathematical expression. You will need to define parameters that are not compartment names in the `parameters` argument

## 4. Multiple transitions from a compartment

In `denim`, transitions between one compartment to multiple compartments are modeled as multinomial transition.

Consider this example:

<center>

```{r, echo=FALSE}
DiagrammeR::grViz("digraph {
  graph [layout = dot, rankdir = LR]
  
  node [shape = rectangle]
  
  S -> I [label = '&#946;SI/N']
  S -> V [label = '5']
  I -> R [label = '0.9 -> d_gamma(1/3, 2)']
  I -> D [label = '0.1 -> d_lognormal(2, 0.5)']
  }",
  width = 500, height = "100%")
```

</center>

There are two scenarios in this example:

-   Susceptible individuals can be infected or vaccinated. The underlying assumption here is 50% of susceptible will be infected (`S -> I`), and the remaining 50% will get vaccinated (`S -> V`).

-   Infected individuals can recover or die. If the mortality probability is known, we can implement it into the model, for example by defining `0.9 * I -> R` (90% individuals will recover) and then `0.1 * I -> D` (10% of them die). By doing so, we ensure that the mortality probability is 10%, while also define the length of stay of individuals at the infected state before recover or die follows gamma or log-normal distribution, respectively.

We can define the model for this example as follows:

```{r, fig.width = 5}
transitions <- list(
  "S -> I" = "beta * S * I / N",
  "S -> V" = 5,
  "0.9 * I -> R" = d_gamma(1/3, 2),
  "0.1 * I -> D" = d_lognormal(2, 0.5)
)

initialValues <- c(
  S = 999, 
  I = 1, 
  R = 0,
  V = 0,
  D = 0
)

parameters <- c(
  beta = 0.12,
  N = 1000
)

simulationDuration <- 10
timeStep <- 0.01

mod <- sim(transitions = transitions, 
           initialValues = initialValues, 
           parameters = parameters, 
           simulationDuration = simulationDuration, 
           timeStep = timeStep)

head(mod)
plot(mod)
```

For a more detailed explanation for multinomial transition, refer to [this article](./multinomial.html)

## 5. Another example

<center>

```{r, echo=FALSE}
DiagrammeR::grViz("digraph {
  graph [layout = dot, rankdir = LR]
  
  node [shape = rectangle]
  
  S -> I [label = '&#946;S(I + IV)/N']
  S -> V [label = '2']
  I -> D [label = '0.1 -> d_lognormal(2, 0.5)']
  I -> R [label = '0.9 -> d_gamma(1/3, 2)']
  V -> IV [label = '0.1 * &#946;V(I + IV)/N']
  IV -> R [label = 'd_exponential(2)']
  }",
  width = 700, height = "100%")
```

</center>

```{r, fig.width = 5}
transitions <- list(
  "S -> I" = "beta * S * (I + IV) / N",
  "S -> V" = 2,
  "0.1 * I -> D" = d_lognormal(2, 0.5),
  "0.9 * I -> R" = d_gamma(rate = 1/3, shape = 2),
  "V -> IV" = "0.1 * beta * V * (I + IV) / N",
  "IV -> R" = d_exponential(2)
)

initialValues <- c(
  S = 999, 
  I = 1, 
  R = 0,
  V = 0,
  IV = 0,
  D = 0
)

parameters <- c(
  beta = 0.12,
  N = 1000
)

simulationDuration <- 10
timeStep <- 0.01

mod <- sim(transitions = transitions, 
           initialValues = initialValues, 
           parameters = parameters, 
           simulationDuration = simulationDuration, 
           timeStep = timeStep)
plot(mod)
```

```{r}
transitions <- list(
  "M -> S" = 100,
  "S -> I" = transprob(1),
  "S -> R" = transprob(1)
)

initialValues <- c(
  M = 1000,
  S = 500, 
  I = 0, 
  R = 0
)

simulationDuration <- 10
timeStep <- 0.01

mod <- sim(transitions = transitions, 
           initialValues = initialValues, 
           parameters = parameters, 
           simulationDuration = simulationDuration, 
           timeStep = timeStep)
head(mod)
```
